The Kapowski Chronicles: A Complete, Volumetric Cortical Thickness Open Source Pipeline with Evaluation on Public Data

### 2012-07-01 KapowskiChronicles/intro.tex
ENH: Brian's edits.

Neuroscientific investigations into cortical morphological changes/differences have illuminated
interesting correlations with normal and pathological neurodevelopment in 
addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
computational methods for modeling the
cortical gray matter have resulted in numerous studies 
exploring such relationships.  
These analysis techniques
have been instrumental in demonstrating 
the presence of regional cortical abnormalities in such conditions as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Further study 
topics exploring correlations with cortical thickness include
normal aging \citep{kochunov2011},
gender differences \citep{luders2006a},
anatomical asymmetries 
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
white collar criminals \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies have demonstrated possible functional 
connectivity relationships using cortical thickness measures
\cite{worsley2005,lerch2006,he2007}.
Although these study findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

Broadly characterized as surface mesh-based or volumetric \citep{scott2009,clarkson2011}, such 
techniques have been introduced and extensively developed over 
the last $10+$ years.  Representative of the former is the
Freesurfer%
\footnote{
http://surfer.nmr.mgh.harvard.edu/
}
cortical modeling software package \citep{dale1999,fischl1999,fischl2000,fischl2002,fischl2004}
which owes its popularity to public availability, excellent documentation, 
good performance, and  integration with other toolkits, such as the extensive FMRIB software 
library (FSL) \citep{smith2004}.  Similar to other surface
approaches (e.g. \cite{davatzikos1996,magnotta1999,macdonald2000,kim2005}), the pial
and white matter surfaces from individual subject MR data are modeled with polygonal meshes  
which are then used to determine local cortical thickness based on a specified correspondence between 
the surface models.

Image volumetric (or meshless) techniques are varied both in algorithmic terms as well as
the underlying definition of cortical thickness.  An early, foundational technique is the 
method of \cite{jones2000} in which the inner and outer surface geometry is used to determine the
solution to Laplace's equation in which thickness is measured by integrating along the 
tangents of the resulting field lines spanning the boundary surfaces.  Subsequent contributions
improved upon the original formulation.  For example, in \cite{yezzi2003}, an Eulerian PDE approach
was proposed to facilitate the computation of correspondence paths.  Extending the surface-based
work of \cite{macdonald2000}, the hybrid approach of
\cite{kim2005} uses the discrete Laplacian field to deform the white matter surface mesh towards the 
pial surface.    Although the Laplacian-based approach has several advantages
including generally lower computational times and 
non-crossing correspondence paths, direct correlative assessments with histology
are difficult as the quantified distances 
are not necessarily Euclidean.  Other volumetric algorithms have employed coupled
level sets \citep{zeng1999}, model-free intelligent search strategies either normal to 
the gray-white matter interface \citep{scott2009} or using a min-max rule \citep{clement-vachet2011}.
Most relevant to this work is the DiReCT (Diffeomorphic Registration-based 
Cortical Thickness) algorithm proposed by \cite{das2009} in which a 
a registration-based approach proposed where the derived diffeomorphic mapping between the 
white and pial matter surfaces is used to propagate thickness values 
through the cortical gray matter.

Despite the numerous techniques that have been proposed in the literature (of which
only a fraction has been cited), several requisite processing components are 
common to many of them which are prior to quantification of cortical thickness.
These include inhomogeneity correction, skull stripping, and $n$-tissue segmentation 
for differentiating the gray and white matters.  For post-processing statistical analysis 
across populations, construction of population-specific unbiased templates
is also  beneficial.
In addition, some of these steps might include a crucial registration component, e.g. 
propagating template-based tissue priors for improved segmentation.  The requisite
additional components coupled with the general lack of availability of published
algorithms \cite{kovacevic2006} inhibits performing studies by external researchers 
and makes comparative evaluations difficult.  For example, one recent evaluation 
study \citep{clarkson2011} compared
Freesurfer (a surface-based method) with two volumetric methods \citep{jones2000,das2009}.
Whereas the entire Freesurfer processing pipeline has been made publicly available, 
tuned by the original authors in terms of implementation, and described in great detail 
(specifically in terms of suggested parameters); both volumetric methods were 
implemented solely by the authors of the evaluation (not the actual software developers) using 
unspecified parameters making the comparisons less than ideal.

In this work we describe our cortical thickness software pipeline which 
we have made open source as part of our Advanced Normalization Tools (ANTs) 
repository.  This includes all the necessary preprocessing steps consisting
of well-vetted previously published algorithms for bias correction \cite{tustison2010},
brain extraction \cite{avants2010a}, $n$-tissue segmentation \cite{avants2011a},
template construction \cite{avants2010}, and image normalization \cite{avants2011}.
We also describe improvements made to the original DiReCT algorithm \cite{das2009}.
Equally as important, we demonstrate how to coordinate
these pipeline components and provide a set of useful parameters
which are used to analyze the publicly available IXI data set. This is 
encapsulated in a well-documented shell script which is also available in ANTS. 
Furthermore, we provide all the derived image data and other processing scripts on 
a github repository specifically meant for this publication.  This permits
the set of results described in this work to be fully reproducible thus
permitting other researchers to use a complete volumetric pipeline for measuring
cortical thickness.

>>>>

% Neuroscientific investigations into cortical morphological
% changes/differences have illuminated interesting correlations with
% normal and pathological neurodevelopment in  addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
imaging-based structural analysis of the brain now plays a fundamental role
in identifying the relationship between cortical morphology, disease and cognition.
Quantitative cortical measures are able to identify abnormalities in such conditions as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Cortical thickness also
varies normally as a function of age \citep{kochunov2011},
gender \citep{luders2006a}, handedness
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
tendency toward criminality \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies demonstrate possible functional 
connectivity relationships using cortical thickness measures
\cite{worsley2005,lerch2006,he2007}.
Although these findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

Computational methods for analyzing the cortex may be 
broadly characterized as surface mesh-based or volumetric \citep{scott2009,clarkson2011}.  Representative of the former is the
Freesurfer%
\footnote{
http://surfer.nmr.mgh.harvard.edu/
}
cortical modeling software package \citep{dale1999,fischl1999,fischl2000,fischl2002,fischl2004}
which owes its popularity to public availability, excellent documentation, 
good performance, and  integration with other toolkits, such as the extensive FMRIB software 
library (FSL) \citep{smith2004}.  Similar to other surface
approaches (e.g. \cite{davatzikos1996,magnotta1999,macdonald2000,kim2005}), the pial
and white matter surfaces from individual subject MR data are modeled with polygonal meshes  
which are then used to determine local cortical thickness based on a specified correspondence between 
the surface models.

Image volumetric (or meshless) techniques vary both in algorithmic terms as well as
the underlying definition of cortical thickness.  An early, foundational technique is the 
method of \cite{jones2000} in which the inner and outer surface geometry is used to determine the
solution to Laplace's equation in which thickness is measured by integrating along the 
tangents of the resulting field lines spanning the boundary surfaces.  Subsequent contributions
improved upon the original formulation.  For example, in \cite{yezzi2003}, an Eulerian PDE approach
was proposed to facilitate the computation of correspondence paths.  Extending the surface-based
work of \cite{macdonald2000}, the hybrid approach of
\cite{kim2005} uses the discrete Laplacian field to deform the white matter surface mesh towards the 
pial surface.    Although the Laplacian-based approach has several advantages
including generally lower computational times and 
non-crossing correspondence paths, direct correlative assessments with histology
are difficult as the quantified distances 
are not necessarily Euclidean.  Other volumetric algorithms employ coupled
level sets \citep{zeng1999}, model-free intelligent search strategies either normal to 
the gray-white matter interface \citep{scott2009} or using a min-max rule \citep{clement-vachet2011}.
Most relevant to this work is the DiReCT (Diffeomorphic Registration-based 
Cortical Thickness) algorithm proposed by \cite{das2009} in which a 
a registration-based approach proposed where the derived diffeomorphic mapping between the 
white and pial matter surfaces is used to propagate thickness values 
through the cortical gray matter.  A unique benefit of DiReCT is that it
naturally estimates the boundaries of buried sulci by employing a
diffeomorphic constraint on the probabilistic estimate of the gray
matter and cerebrospinal fluid interface.

Despite the variety of techniques for estimating cortical thickness
from imaging data (of which
only a fraction are cited), several common pre-processing components
may be identified.
The most fundamental of these include inhomogeneity correction, skull stripping, and $n$-tissue segmentation 
for differentiating the gray and white matter.  For statistical analysis 
across large populations, construction of population-specific unbiased templates
is also beneficial \cite{}.
In addition, some of these steps might include a crucial registration component, e.g. 
propagating template-based tissue priors for improved segmentation.  The requisite
additional components coupled with the general lack of availability of published
algorithms \cite{kovacevic2006} inhibits performing studies by external researchers 
and makes comparative evaluations difficult.  For example, one recent evaluation 
study \citep{clarkson2011} compared
Freesurfer (a surface-based method) with two volumetric methods \citep{jones2000,das2009}.
Whereas the entire Freesurfer processing pipeline has been made publicly available, 
tuned by the original authors in terms of implementation, and described in great detail 
(specifically in terms of suggested parameters); both volumetric methods were 
implemented solely by the authors of the evaluation (not the actual software developers) using 
unspecified parameters making the comparisons less than
ideal. \textcolor{blue}{might also note the Klein 2011 evaluation that
  raises the difficulty of comparing volumetric and surface-based
  methods and the bias induced by the domain representation.}

In this work, we describe our C$++$ based cortical thickness pipeline
which is freely available as part of the Advanced Normalization Tools
(ANTs) software package.  This includes all the necessary preprocessing steps consisting
of well-vetted previously published algorithms for bias correction \cite{tustison2010},
brain extraction \cite{avants2010a}, $n$-tissue segmentation \cite{avants2011a},
template construction \cite{avants2010}, and image normalization \cite{avants2011}.
We also describe improvements made to the original DiReCT algorithm \cite{das2009}.
Equally as important, we demonstrate how to coordinate
these pipeline components and provide a set of useful parameters
which are employed to analyze the publicly available IXI data set. The
full pipeline and parameter set is 
encapsulated in a well-documented shell script which is also available in ANTs. 
Furthermore, we provide all the derived image data and other processing scripts on 
a github repository specifically meant for this publication.  The
availability of both the code and data permits
the set of results described in this work to be fully reproducible.  This
permits other researchers to contrast their own results against
this baseline processing and to adapt the given volumetric pipeline for measuring
cortical thickness in their own datasets.

### 2012-07-01 KapowskiChronicles/methods.tex
ENH: Brian's edits.

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternating between prior-based segmentation and white matter posterior
        probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

@@

Normalizing images to a standard coordinate system
reduces intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as the selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \citep{Talairach1988}, or a publicly available averaged group of
subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
templates.  Additionally, mean templates constructed from labeled
data can be used to construct spatial priors for improving segmentation
algorithms.
The work of \cite{avants2010} explicitly models the geometric component of the 
normalized space during optimization to produce such mean templates.  Coupling the intrinsic symmetry of 
SyN pairwise registration \citep{avants2011} and an
optimized sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \citep{avants2010} with arbitrary similarity metric choice.

@@

@@

  The warped template probability map is thresholded at 0.5 and the resulting mask is dilated
with a radius of 2.  Atropos is used to generate an initial 3-tissue segmentation estimate within the mask
region.  Each of the three tissue masks undergo specific morphological operations which are then
combined to create a brain extraction mask for use in the rest of the cortical thickness workflow.

A comparison using open access brain data with publicly available brain extraction algorithms 
including  AFNI's \verb#3dIntracranial# \citep{ward1999}, FSL's \verb#BET2# \citep{smith2002}, 
Freesurfer's \verb#mri_watershed# \citep{segonne2004}, and BrainSuite \citep{dogdas2005}  demonstrated
that our combined registration/segmentation approach \citep{avants2010a} performs at the top level
alongside BrainSuite (tuned) and FreeSurfer.

@@

Although the basic formulation of DiReCT as reported in this work is as it 
was introduced in \cite{das2009}, we have made several 
improvements (perhaps the most significant being that this particular 
ITK-compatible implementation has been significantly multi-threaded,
is written in ITK coding style, and has been made publicly available through 
ANTs complete with a unique user interface design developed specifically for 
ANTs tools.

>>>>

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
    \textcolor{blue}{do you usually bias correct the raw image?  phil
      had some problems with the automated (otsu) masking employed
      therein .... esp with elderly brain}
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternating between prior-based segmentation and white matter posterior
        probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

@@

Normalizing images to a standard coordinate system
reduces intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as the selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \citep{Talairach1988}, or a publicly available averaged group of
subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
templates.  Additionally, mean templates constructed from labeled
data can be used to construct spatial priors for improving segmentation
algorithms.
The work of \cite{avants2010} explicitly models the geometric component of the 
normalized space during optimization to produce such mean templates.  Coupling the intrinsic symmetry of 
SyN pairwise registration \citep{avants2011} and an
optimized shape-based sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \citep{avants2010} with arbitrary similarity metric choice.

@@

\textcolor{blue}{what is the motivation of the multivar template here?
  it would seem useful to add an FA component to the segmentation step
  to aid in cortical / wm delineation which would motivate the
  multivar template approach but might be difficult to establish that
  this actually helps .... though we could compare both ways : univar
  and multivar }

@@

  The warped template probability map is thresholded at 0.5 and the resulting mask is dilated
with a radius of 2.  Atropos is used to generate an initial 3-tissue segmentation estimate within the mask
region.  Each of the three tissue masks undergo specific morphological operations which are then
combined to create a brain extraction mask for use in the rest of the
cortical thickness workflow.  \textcolor{blue}{does there need to be a
  bit more technical detail here?  while you can refer to the script,
  why perform these operations?  there are numerous references, most
  germane probably some recent stuff from J Prince ( i think ) and the
  freesurfer watershed approach which came from ... i forget ... maybe
  one of the french groups.}

A comparison using open access brain data with publicly available
brain extraction algorithms including AFNI's \verb#3dIntracranial#
\citep{ward1999}, FSL's \verb#BET2# \citep{smith2002}, Freesurfer's
\verb#mri_watershed# \citep{segonne2004}, and BrainSuite
\citep{dogdas2005} demonstrated that our combined
registration/segmentation approach \citep{avants2010a} performs at the
top level alongside BrainSuite (tuned) and FreeSurfer.
\textcolor{blue}{ok you have the segonne ref here ... }

@@

Although the basic formulation of DiReCT as reported in this work is as it 
was introduced in \cite{das2009}, we have made several 
improvements.  Perhaps the most significant advance is that this particular 
ITK-compatible implementation has been significantly multi-threaded,
is written in ITK coding style, and has been made publicly available through 
ANTs complete with a unique user interface design developed specifically for 
ANTs tools.

### 2012-07-01 KapowskiChronicles/paper.tex
ENH: Brian's edits.

\author[label1]{Nicholas J.~Tustison$^\dagger$
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.\\ \\
  $\dagger$The first two authors contributed equally to this work.}
\author[label2]{Brian B.~Avants$^\dagger$}
\author[label2]{Philip A.~Cook}
\author[label2]{Gang Song}
\author[label2]{Sandhitsu R.~Das}
\author[label2]{Jeffrey R.~Duda}
\author[label3]{Niels van Strien}
\author[label1]{James R.~Stone}
\author[label2]{James C.~Gee}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Circulation and Medical Imaging, Norwegian University of Science and Technology, Trondheim, Norway}

@@

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function, and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools
(ANTS) cortical thickness automated pipeline
comprising well-vetted components  such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias
correction),
Atropos ($n$-tissue segmentation), and DiReCT (cortical thickness) 
all developed as part of the ANTS open science effort.  
Complementing the open source aspect of ANTS we confirm 
previous findings of
gender differences, age correlation, and hemispherical asymmetry 
in cortical thickness using the publicly available IXI data
consisting of multimodal structural MRI of approximately 600 subjects.
To further promote open science and use of the proposed tools,
 all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}

>>>>

\author[label1]{Nicholas J.~Tustison$^\dagger$
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.\\ \\
  $\dagger$The first two authors contributed equally to this work.}
\author[label2]{Brian B.~Avants$^\dagger$}
\author[label2]{Philip A.~Cook}
\author[label2]{Gang Song}
\author[label2]{Sandhitsu R.~Das}
\author[label2]{Jeffrey R.~Duda}
\author[label3]{Niels van Strien}
\author[label1]{James R.~Stone}
\author[label2]{James C.~Gee}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Circulation and Medical Imaging,
  Norwegian University of Science and Technology, Trondheim,
  Norway---\textcolor{blue}{should be netherlands not sure of specifics}}

@@

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function, and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools
(ANT\textcolor{blue}{s}) cortical thickness automated pipeline
comprising well-vetted components such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias
correction),
Atropos ($n$-tissue segmentation), and DiReCT (cortical thickness) 
all developed as part of the ANT\textcolor{blue}{s} open science effort.  
Complementing the open source aspect of ANT\textcolor{blue}{s} we confirm 
previous findings of
gender differences, age correlation, and hemispherical asymmetry 
in cortical thickness using the publicly available IXI data
consisting of multimodal structural MRI of approximately 600 subjects.
To further promote open science and use of the proposed tools,
 all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}
